Mechanising Hylomorphisms for Extraction of Executable Code

### 2025-03-21 paper/paper.tex
Rewrites to conclusions and related work

Tesson et al.\ demonstrated the efficacy of leveraging Coq to establish an
approach for implementing a robust system dedicated to verifying the correctness
of program transformations for functions that manipulate
lists~\cite{10.1007/978-3-642-17796-5_10}. Murata and Emoto went further and
formalised recursion schemes in Coq~\cite{MurataE19}. Their development does not
include hylomorphisms and dynamorphisms, and relies on the functional
extensionality axiom, as well as further extensionality axioms for each
coinductive datatype that they use. They do not discuss the extracted code
from their formalisation. Larchey-Wendling and Monin encode recursion schemes in
Coq, by formalising computational graphs of algorithms~\cite{larchey2022braga}.
Their work does not focus on encoding generic recursion schemes, and proving
their algebraic laws. Castro-Perez et al.~\cite{farmsCastro} encode the laws of
hylomorphisms as part of the type system of a functional language to calculate
parallel programs from specifications. Their work focuses on parallelism, and
they do not formalise their approach in a proof assistant, and the laws of
hylomorphisms are axioms in their system.

% \textbf{Fusion Optimisation}.
% %
% \label{sec:fusion}
Fusion in functional programming is a more general technique for combining multiple
computations into a single pass to eliminate intermediate data structures or
repeated traversals, improving efficiency. The fusion we have discussed in this
paper relating to program calculation is one form of fusion, but there are other
approaches, and prior research has explored the mechanisation and verification
of fusion transformations.

Fusion has been applied to recursive tree traversals, in order to reduce the
number of times each section of the tree must be visited. Research has
demonstrated that this sort of fusion can be performed automatically by a
compiler~\cite{10.1145/3133900}, and the correctness of tree traversal fusion
has been verified mechanically in Coq~\cite{10.1145/3377555.3377884}.

@@

%\mvol{Need more general citations for program calculation.}
% \textbf{Divide and Conquer Recursion}.%
% \label{sec:reccoq}
Abreu et al.~\cite{AbreuDHJMS23} encode divide-and-conquer computations in Coq,
using a recursion scheme in which termination is entirely enforced by its
typing. This is a significant advance, since it avoids \emph{completely} the
need for termination proofs. Their work differs from ours in that they require
the functional extensionality axiom, and the use of impredicative \coq{Set}. The
authors justify well the use of impredicative \coq{Set} and its compatibility
with the functional extensionality axiom. In contrast, our development remains
entirely \emph{axiom-free}, and we retain the ability to extract natural-looking
OCaml code.
% Another key difference with our
% approach is that they do not discuss what the resulting extracted code
% \emph{looks like}.
% (that is, whether the extracted OCaml code resembles the
% natural formulation of the recursive function).
Through experiments, we found that code extracted from their formalisation makes
heavy use of unsafe casts and has a generally complex structure that may be
hard to understand and integrate with other code. Due to the great benefit
of entirely avoiding termination proofs, it would be interesting to extend their
approach to improve code extraction.

%\textbf{Termination Checking of Nonstructural Recursion}.% \label{sec:termcheck}
% Termination and productivity are non-trivial properties of programs, and various
% methods have been proposed for checking that these properties hold. A common
% approach is guardedness checking~\cite{Eduardo94}, which is a
% \emph{syntactic} check that definitions avoid the introduction of
% non-normalizable terms. This sort of check generally looks for a
% \emph{structural decrease} of arguments in the recursive calls of a function.
% Coq uses such a check, and it works for many classic functional programming
% patterns (like \haskell{map} and \haskell{foldr}). However, some desirable
% definitions will not be accepted by such checks.
%In particular,
The problem of \emph{nonstructural recursion} (including
divide-and-conquer algorithms) is well-studied~\cite{BoveKS16}.
% A distinction between standard and
% nonstandard recursion lies in the fact that standard recursion adheres strictly
% to recursing on pattern variables, avoiding reliance on results from other
% function calls. Nonstandard recursion may deviate from this convention, allowing
% for more flexibility.
Certain functions that are not structurally recursive can be reformulated using
a nonstandard approach to achieve structural recursion~\cite{AbreuDHJMS23}.
Take, for instance, division by iterated subtraction, which is inherently
non-structurally recursive since it involves recursion based on the result of a
subtraction. There is a nonstandard implementation of division found on Coq's
standard library, which involves a four-argument function that effectively
combines subtraction and division. Similarly, the mergesort in Coq's standard
library uses an ``explicit stack of pending merges'' in order to avoid issues
with nonstructural definitions. There is a major downside, however; as noted by
Abreu et al., the result is ``barely recognisable as a form of
mergesort''~\cite{AbreuDHJMS23}.
There are common approaches in Coq to deal with termination, but none of these
approaches address program calculation techniques, and the mechanisation of
fusion laws.

@@

\section{Conclusions and Future Work}
In this work we mechanise hylomorphisms and their algebraic laws and use them
to perform optimisations for programs than can be later extracted into human-readable and runnable code.
We do this in a \emph{fully
  axiom-free}.

This formalisation allows the use of program calculation techniques in Coq to derive
formally optimised implementations from naive specifications. Furthermore, the
rewritings that are applied to specifications are formal, machine-checked
proofs that the resulting program is extensionally equal to the input
specification.

As part of the future improvements, we will study how to mitigate the problem of
setoids. At the moment, we use a short ad-hoc tactic that is able to
automatically discharge many of these proofs in simple settings.  We will study
the more thorough and systematic use of proof automation for respectful
morphisms. Generalised rewriting in proofs involving setoids tends to be quite
slow, due to the large size of the terms that need to be rewritten. Sometimes,
this size is hidden in implicit arguments and coercions. We will study
alternative formulations to try to improve the performance of the rewriting
tactics (e.g.\ canonical structures).  Currently, Coq is unable to inline a
number of trivially inlineable definitions.  We will study alternative
definitions, or extensions to Coq's code extraction mechanisms to force the
full inlining of all container code that is used in hylomorphisms.  Finally,
proving termination still remains a hurdle. In our framework this reduces to
proving that the anamorphism
terminates in all inputs, and we provide a convenient connection to
well-founded recursion. Furthermore, recursive coalgebras compose with natural
transformations, which allows the reuse
of a number of core recursive coalgebras. A possible interesting future line of work
is the use of the approach by Abreu et al.~\cite{AbreuDHJMS23} in combination with ours to
improve code extraction from divide-and-conquer computations whose termination
does not require an external proof.

>>>>

There is existing prior work on formalising recursion schemes in a proof
assistant.  Tesson et al.\ demonstrated the efficacy of leveraging Coq to
establish an approach for implementing a robust system dedicated to verifying
the correctness of program transformations for functions that manipulate
lists~\cite{10.1007/978-3-642-17796-5_10}. Murata and Emoto went further and
formalised recursion schemes in Coq~\cite{MurataE19}. Their development does not
include hylomorphisms and dynamorphisms, and relies on the functional
extensionality axiom, as well as further extensionality axioms for each
coinductive datatype that they use. They do not discuss the extracted code from
their formalisation. Larchey-Wendling and Monin encode recursion schemes in Coq,
by formalising computational graphs of algorithms~\cite{larchey2022braga}.
Their work does not focus on encoding higher-order generic recursion schemes,
and proving their algebraic laws. Castro-Perez et al.~\cite{farmsCastro} encode
the laws of hylomorphisms as part of a type system to calculate parallel
programs from specifications. Their work focuses on parallelism, and they do not
formalise their approach in a proof assistant, and they treat the  laws of
hylomorphisms as axioms.

%\mvol{Need more general citations for program calculation.}
% \textbf{Divide and Conquer Recursion}.%
% \label{sec:reccoq}
Abreu et al.~\cite{AbreuDHJMS23} encode divide-and-conquer computations in Coq,
using a recursion scheme in which termination is entirely enforced by its
typing. This is a significant advance, since it avoids \emph{completely} the
need for termination proofs. Their work differs from ours in that they require
the functional extensionality axiom, and the use of impredicative \coq{Set}. The
authors justify well the use of impredicative \coq{Set} and its compatibility
with the functional extensionality axiom. In contrast, our development remains
entirely \emph{axiom-free}, and we retain the ability to extract natural-looking
OCaml code.
% Another key difference with our
% approach is that they do not discuss what the resulting extracted code
% \emph{looks like}.
% (that is, whether the extracted OCaml code resembles the
% natural formulation of the recursive function).
Through experiments, we found that code extracted from their formalisation makes
heavy use of unsafe casts and has a generally complex structure that may be hard
to understand and integrate with other external code. Due to the great benefit
of entirely avoiding termination proofs, it would be interesting to extend their
approach to improve code extraction.

The problem of \emph{nonstructural recursion} (including
divide-and-conquer algorithms) is well-studied~\cite{BoveKS16}.
% A distinction between standard and
% nonstandard recursion lies in the fact that standard recursion adheres strictly
% to recursing on pattern variables, avoiding reliance on results from other
% function calls. Nonstandard recursion may deviate from this convention, allowing
% for more flexibility.
Certain functions that are not structurally recursive can be reformulated using
a nonstandard approach to achieve structural recursion.  For exammple, the
mergesort in Coq's standard library uses an ``explicit stack of pending merges''
in order to avoid issues with nonstructural definitions. There is a major
downside, however; as noted by Abreu et al., the result is ``barely recognisable
as a form of mergesort''~\cite{AbreuDHJMS23}.  There are common approaches in
Coq to deal with termination~\cite{DBLP:journals/pacmpl/SozeauM19,
  DBLP:conf/icfp/Sozeau07}, but none of these approaches address program
calculation techniques, and the mechanisation of fusion laws.

% \textbf{Fusion Optimisation}.
% %
% \label{sec:fusion}
The fusion laws that we formalise in this paper are applied in the more general
context of compilers for functional programming lanmguages.  Fusion has been
applied to recursive tree traversals, in order to reduce the number of times
each section of the tree must be visited. Research has demonstrated that this
sort of fusion can be performed automatically by a
compiler~\cite{10.1145/3133900}. The correctness of tree traversal fusion has
been verified mechanically in Coq~\cite{10.1145/3377555.3377884}, but the
authors do not formalise unfolds and hylomorphisms.

@@

%\textbf{Termination Checking of Nonstructural Recursion}.% \label{sec:termcheck}
% Termination and productivity are non-trivial properties of programs, and various
% methods have been proposed for checking that these properties hold. A common
% approach is guardedness checking~\cite{Eduardo94}, which is a
% \emph{syntactic} check that definitions avoid the introduction of
% non-normalizable terms. This sort of check generally looks for a
% \emph{structural decrease} of arguments in the recursive calls of a function.
% Coq uses such a check, and it works for many classic functional programming
% patterns (like \haskell{map} and \haskell{foldr}). However, some desirable
% definitions will not be accepted by such checks.
%In particular,

@@

\section{Conclusions and Future Work}
In this work we mechanise hylomorphisms and their algebraic laws and use them to
perform program calculation and optimisations. The resulting programs can be
later extracted into idiomatic and runnable code. Our mechanisation is
\emph{fully axiom-free}. This formalisation allows us to use Coq as a framework
for program calculation in a way that is close to common practice in pen and
paper program calculation proofs. This implies that every rewriting step is the
result of applying a formal, machine-checked proof that the resulting program is
extensionally equal to the input specification.

As part of the future improvements, we will study how to mitigate the problem of
setoids. At the moment, we use a short ad-hoc tactic that is able to
automatically discharge many of these proofs in simple settings.  We will study
the more thorough and systematic use of proof automation for proper morphisms.
Generalised rewriting in proofs involving setoids tends to be quite slow, due to
the large size of the terms that need to be rewritten. Sometimes, this size is
hidden in implicit arguments and coercions. We will study alternative
formulations to try to improve the performance of the rewriting tactics
(e.g.\ canonical structures).  Currently, Coq is unable to inline a number of
trivially inlineable definitions.  We will study alternative definitions, or
extensions to Coq's code extraction mechanisms to force the full inlining of all
container code that is used in hylomorphisms.  Finally, proving termination
still remains a hurdle. In our framework this reduces to proving that the
anamorphism terminates in all inputs, and we provide a convenient connection to
well-founded recursion. Furthermore, recursive coalgebras compose with natural
transformations, which allows the reuse of a number of core recursive
coalgebras. A possible interesting future line of work is the use of the
approach by Abreu et al.~\cite{AbreuDHJMS23} in combination with ours to improve
code extraction from divide-and-conquer computations whose termination does not
require an external proof.
